---
title: "CMPINF 2130 Summer 2022 - Week 01"
subtitle: "Our first ggplot graphic"
author: "Dr. Joseph P. Yurko"
date: '2022-05-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_packages}
library(tidyverse)
```


## Make data

```{r, make_01}
df_a <- data.frame(x = seq(-2, 2, length.out = 9))
```


```{r, make_02}
df_a
```

Filter or conditionally subset all rows where `x < 0`.  

```{r, make_03}
filter(df_a, x < 0)
```

The forward pipe operator allows us to pass objects into functions "left-to-right".  

```{r, make_04}
df_a %>% filter(x < 0)
```

Chain together multiple data manipulation steps those steps will read left-to-right.  

```{r, make_05}
df_a %>% filter(x < 0) %>% dim()
```

```{r, make_06}
dim( filter(df_a, x<0) )
```

```{r}
df_a %>% 
  mutate(y = x^2)
```

```{r}
df_a
```

```{r}
mutate(df_a, y = x^2, z = 3 * x, w = y %% 1)
```

```{r, anamed_for_the_chunk}
df_a %>% 
  mutate(y = x^2,
         z = 3 * x,
         w = y %% 1) %>% 
  filter(z < 0) %>% 
  rename(x1 = x, yy = y)
```



## Visualize data


```{r, viz_01}
df_a %>% 
  mutate(y = x^2) %>% 
  ggplot()
```


```{r}
df_a %>% 
  mutate(y = x^2) %>% 
  ggplot(mapping = aes(x = x, y = y))
```

```{r}
df_a %>% 
  mutate(y = x^2) %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(size = 8, color = 'blue', shape = 15, alpha = 0.55)
```

```{r}
df_a %>% 
  mutate(y = x^2) %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_line(size = 2.5) +
  geom_point(size = 8, color = 'blue', shape = 15, alpha = 0.33)
```












































