---
title: "CMPINF 2130 Summer 2022 - Week 02"
subtitle: "aesthetic deep dive"
author: "Dr. Joseph P. Yurko"
date: '2022-05-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_packages}
library(tidyverse)
```

## Anscombe

```{r, a_1}
anscombe %>% glimpse()
```

```{r, a_2}
anscombe
```

Start working with the first dataset within Anscombe.  

```{r, a_3}
anscombe %>% 
  select(x1, y1)
```

```{r, a_4}
anscombe %>% 
  select(x = x1, y = y1)
```

```{r, a_5}
anscombe %>% 
  rename(x = x1, y = y1)
```

Or we can exploit common naming patterns with `dplyr` helper functions.  

```{r, a_6}
anscombe %>% 
  select(ends_with('1'))
```

Let's combine the helper with the `rename()` function.  

```{r, a_7}
a1 <- anscombe %>% 
  select(ends_with('1')) %>% 
  rename(x = x1, y = y1)
```

## Visualize the a1 data set

The 3 major components of any (statistical) figure.  

* data - a table (dataframe)  
* aesthetics - what do we want to see (the columns from a table)  
* geometric objects - "how" do we want to see the variables  

Create a scatter plot to visualize the relationship between `y` and `x`.  

```{r, v_1}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point()
```



Maybe you mapped the aesthetics within the `geom_point()` layer...  

```{r, v_2}
a1 %>% 
  ggplot() +
  geom_point(mapping = aes(x = x, y = y))
```



But...what happens if we did NOT map the `x` and `y` aesthetics?  

The code below produces an error and is NOT executed or evaluated.  

```{r, v_3, eval=FALSE}
a1 %>%
  ggplot() +
  geom_point()
```


Some more errors that can occur. Map the `x` aesthetic but forget the `y` aesthetic.  

```{r, v_4}
### the code below produces an error and is commented out
# a1 %>% 
#   ggplot(mapping = aes(x = x)) +
#   geom_point()
```

Same thing with the `y` aesthetic.  

```{r, v_5}
### ERROR!!!
# a1 %>% 
#   ggplot(mapping = aes(y = x)) +
#   geom_point()
```


What happens if we forget about the `aes()` function?  

We need to include the aesthetics within the `aes()` function to instruct ggplot2 to "look" at the provided data. Without `aes()` it cannot find the variable.  

The code chunk below results in an error.  

```{r, v_6, eval=FALSE}
a1 %>% 
  ggplot() +
  geom_point(x = x, y = y)
```


Let's create a data set for the second Anscombe quarter data.  

```{r, v_7}
a2 <- anscombe %>% 
  select(ends_with('2')) %>% 
  rename(x = x2, y = y2)
```


Create a scatter plot between `y` and `x` for the second quartet.  

```{r, v_8}
a2 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point()
```

How does `aes()` handle the other aesthetics.  

Let's make a scatterplot for the `a1` data set again but with blue markers.  
First, try setting the aesthetic, `color` WITHIN `aes()`.  

```{r, v_9}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = 'blue'))
```


Assign the `color` outside of `aes()` within `geom_point()`.  

```{r, v_10}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(color = 'blue')
```



The `aes()` function wants to link or MAP or associate variable values to visual properties.  

```{r, v_11}
a1 %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = 'the value of a variable that does not exist'))
```


If you want to HARD CODE a visual property do NOT use `aes()`.  

Let's map aesthetics to "manually defined variable values" to let us directly compare quartets within Anscombe.  

```{r, x_1}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1)) +
  geom_point(mapping = aes(x = x2, y = y2))
```

Distinguish one data set from the other with color. First, do so by hard coding the colors outside `aes()`.  

```{r, x_2}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1),
             color = 'red', size = 4) +
  geom_point(mapping = aes(x = x2, y = y2),
             color = 'dodgerblue', size = 4)
```

Instead of hard coding we can define the color within `aes()` by "manually mapping to a variable".  

```{r, x_3}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first'),
             size = 4) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second'),
             size = 4)
```


Change the legend guide LABEL with the `scale_` family functions.  


```{r, x_4}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first'),
             size = 4) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second'),
             size = 4) +
  scale_color_discrete('dataset')
```


We can use a different categorical color scale if we'd like.  

```{r, x_5}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first'),
             size = 4) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second'),
             size = 4) +
  scale_color_brewer('dataset', palette = 'Set1')
```

We know the axis labels are not correct, so we can override with the `labs()` function.  

```{r, x_6}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first'),
             size = 4) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second'),
             size = 4) +
  scale_color_brewer('dataset', palette = 'Set1') +
  labs(x = 'x', y = 'y')
```

Also use the marker shape to denote the dataset. Include the shape scale function so ggplot2 knows the variable being represented is the same as color.  

```{r, x_7}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first',
                           shape = 'first'),
             size = 4) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second',
                           shape = 'second'),
             size = 4) +
  scale_color_brewer('dataset', palette = 'Set1') +
  scale_shape_discrete('dataset') +
  labs(x = 'x', y = 'y')
```


Instead of hard coding the marker size, link the marker size to a variable.  

```{r, x_8}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first',
                           shape = 'first',
                           size = 'first')) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second',
                           shape = 'second',
                           size = 'second')) +
  scale_color_brewer('dataset', palette = 'Set1') +
  scale_shape_discrete('dataset') +
  labs(x = 'x', y = 'y')
```


When we want to use `size` with a categorical we need to use a MANAUL scale.  

```{r, x_9}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first',
                           shape = 'first',
                           size = 'first')) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second',
                           shape = 'second',
                           size = 'second')) +
  scale_color_brewer('dataset', palette = 'Set1') +
  scale_shape_discrete('dataset') +
  scale_size_manual('dataset',
                    values = c("first" = 5.5,
                               "second" = 2.5)) +
  labs(x = 'x', y = 'y')
```

We can also modify the marker transparency with the `alpha` aesthetic.  

```{r, x_10}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first',
                           shape = 'first',
                           size = 'first',
                           alpha = 'first')) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second',
                           shape = 'second',
                           size = 'second',
                           alpha = 'second')) +
  scale_color_brewer('dataset', palette = 'Set1') +
  scale_shape_discrete('dataset') +
  scale_size_manual('dataset',
                    values = c("first" = 5.5,
                               "second" = 2.5)) +
  labs(x = 'x', y = 'y')
```



We can use a manual scale to associate transpreancy with a categorical variable.  

```{r, x_11}
anscombe %>% 
  ggplot() +
  geom_point(mapping = aes(x = x1, y = y1,
                           color = 'first',
                           shape = 'first',
                           size = 'first',
                           alpha = 'first')) +
  geom_point(mapping = aes(x = x2, y = y2,
                           color = 'second',
                           shape = 'second',
                           size = 'second',
                           alpha = 'second')) +
  scale_color_brewer('dataset', palette = 'Set1') +
  scale_shape_discrete('dataset') +
  scale_size_manual('dataset',
                    values = c("first" = 5.5,
                               "second" = 2.5)) +
  scale_alpha_manual('dataset',
                     values = c('first' = 0.45,
                                'second' = 0.95)) +
  labs(x = 'x', y = 'y')
```


## 4 data sets

The 4 data sets are constructed to consider the **linear relationship** between `y` and `x`.  

Let's fit a linear model aka "the best fit line" between the output and input for each data set.  

```{r, m_1}
mod_1 <- lm(y ~ x, data = a1)
```

```{r, m_2}
mod_1 %>% summary()
```

What are the coefficients? The intercept and slope?  

```{r, m_3}
coef(mod_1)
```

```{r, m_4}
mod_1 %>% broom::tidy()
```


Fit a linear model for the second data set.  

```{r, m_5}
mod_2 <- lm(y ~ x, data = a2)

mod_2 %>% broom::tidy()
```


Look at the R-squared.  

```{r, m_6}
mod_1 %>% broom::glance()
```

```{r, m_7}
mod_2 %>% broom::glance()
```


## Visualize all 4 datasets

Tidy the `anscombe` data set.  

```{r, tt_1}
a_tidy <- anscombe %>% 
  tibble::rowid_to_column() %>% 
  pivot_longer(cols = !c("rowid")) %>% 
  tidyr::separate(name,
                  c("variable", "dataset"),
                  sep = 1) %>% 
  pivot_wider(names_from = 'variable', values_from = 'value') %>% 
  arrange(dataset)
```

The tidy data long format data are consistent with the data from Seaborn.  

```{r, tt_2}
a_tidy %>% glimpse()
```

Use facets to show the different groups.  

```{r, tt_3}
a_tidy %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point() +
  facet_wrap(~dataset, labeller = 'label_both')
```

We can easily map any aesthetic to the `dataset` column.  

```{r, tt_4}
a_tidy %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = dataset,
                           shape = dataset),
             size = 3.5) +
  facet_wrap(~dataset, labeller = 'label_both') +
  theme(legend.position = 'none')
```

We can easily add in the best fit line with another layer.  

```{r, tt_5}
a_tidy %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = dataset,
                           shape = dataset),
             size = 3.5) +
  geom_smooth(method = lm,
              formula = y ~ x,
              mapping = aes(color = dataset,
                            fill = dataset)) +
  facet_wrap(~dataset, labeller = 'label_both') +
  theme(legend.position = 'none')
```






