---
title: "CMPINF 2130 Summer 2022 - Week 03"
subtitle: "Explore iris"
author: "Dr. Joseph P. Yurko"
date: '2022-05-31'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, load_packages}
library(tidyverse)
```

## Iris

`iris` is part of base R.  

```{r, iris_1}
iris %>% glimpse()
```

## Scatter plot

### 2 variables


```{r, iris_2}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()
```

Manually change the marker size to make it easier to see the plot.  

Change the `fig.` chunk options to modify the figure size as rendered in the report.  

```{r, iris_3, fig.width=12}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5)
```


We can modify the transparency to help see "overlap".  

```{r, iris_4}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, alpha = 0.5)
```

What is the general relationship between these two variables?  

We can add a smoothing trend to help us "see a general pattern through the noise".  

```{r, iris_5}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, alpha = 0.5) +
  geom_smooth(formula = y ~ x, 
              method = lm)
```

### 3 variables

#### Categorical

We can use the marker shape, marker color, marker size, marker transparency as visual attributes to display the 3rd variable.  

However, it's usually best practice to use color and/or shape for categorical variables.  

Let's start by mapping the `Species` to the marker `color` aesthetic.  

```{r, iris_6}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species),
             size = 3.5)
```

We could map all aesthetics in the parent ggplot() call, but I don't like to do this.  

```{r, iris_7}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width,
                       color = Species)) +
  geom_point(size = 3.5)
```

We can try using shape instead of color for categorical variables.  

```{r, iris_8}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5,
             mapping = aes(shape = Species))
```

Sometimes non-solid markers can also help with identifying overlap.  

```{r, iris_9}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5,
             mapping = aes(shape = Species)) +
  scale_shape_discrete(solid = FALSE)
```


When we are only working with 3 variables we can use shape and color for a categorical.  

```{r, iris_10}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5,
             mapping = aes(shape = Species,
                           color = Species))
```

Lastly, let's hard code the transparency to help with overlap.  

```{r, iris_11}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5,
             mapping = aes(shape = Species,
                           color = Species),
             alpha = 0.5)
```

We can manually define the shapes.  

```{r, iris_12}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5,
             mapping = aes(shape = Species,
                           color = Species),
             alpha = 0.5) +
  scale_shape_manual(values = c('setosa' = 20,
                                'versicolor' = 21,
                                'virginica' = 4))
```

Sometimes we just want to look a manually identified BINARIZED version of a categorical variable.  

We can define a new variable with `mutate()` to correspond to the binarized "event".  

```{r, iris_13}
iris %>% 
  mutate(species_binary = ifelse(Species == 'setosa', 
                                 'setosa species',
                                 'something else')) %>% 
  select(Species, species_binary) %>% 
  count(Species, species_binary)
```

We can use our manually created variable to color.  

```{r, iris_14}
iris %>% 
  mutate(species_binary = ifelse(Species == 'setosa',
                                 'setosa species',
                                 'something else')) %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = species_binary),
             size = 3.5, alpha = 0.5)
```


We can manually create variables in `ggplot2` functions.  


```{r, iris_15}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species == 'setosa'),
             size = 3.5, alpha = 0.5)
```

If the legend is ever too long...we can move it to the top or bottom of the figure.  

```{r, iris_16}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species == 'setosa'),
             size = 3.5, alpha = 0.5) +
  theme(legend.position = 'top')
```

We can apply the direct conditional test to multiple aesthetics.  

```{r, iris_17}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species == 'setosa',
                           shape = Species == 'setosa'),
             size = 3.5, alpha = 0.5) +
  theme(legend.position = 'top')
```

To demonstrate working with a larger number of categories let's randomly assign letters to the rows of `iris`.  

```{r, iris_18}
LETTERS
```

Randomly sample or shuffle the letters via the `sample()` function.  

```{r, iris_19}
set.seed(2130)
sample(LETTERS)
```

We need to sample **WITH REPLACEMENT**.  

```{r, iris_20}
set.seed(2130)
sample(LETTERS, size = nrow(iris), replace = TRUE)
```

Let's add the randomly shuffled letters to `iris`.  

```{r, iris_21}
set.seed(2130)
iris %>% 
  mutate(random_letters = sample(LETTERS, size = n(), replace = TRUE)) %>% 
  count(random_letters)
```


First, check adding a column based on a condition.  

```{r, iris_22}
iris %>% 
  mutate(a_new_variable = 'A',
         b_new_variable = Species == 'setosa',
         c_new_variable = ifelse(Species == 'setosa',
                                 'the setosa species',
                                 'a different species'),
         d_new_variable = ifelse(Species == 'setosa',
                                 101,
                                 -101)) %>% 
  head()
```


Now that we know how to add the random shuffled letter colum let's color by this new variable.  

```{r, iris_23}
set.seed(2130)
iris %>% 
  mutate(random_letters = sample(LETTERS, size = n(), replace = TRUE)) %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = random_letters),
             size = 3.5)
```

What if we use shape as an aesthetic for our randomly shuffled letters?  

```{r, iris_24}
set.seed(2130)
iris %>% 
  mutate(random_letters = sample(LETTERS, size = n(), replace = TRUE)) %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = random_letters,
                           shape = random_letters),
             size = 3.5)
```


Other categorical color scales.  

Modify the luminance or lightness of the markers.  

```{r, iris_25}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, 
             mapping = aes(color = Species)) +
  scale_color_discrete(l = 95)
```

Modifying the colors may require us to change the background panel colors.  

```{r, iris_26}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, 
             mapping = aes(color = Species)) +
  scale_color_discrete(l = 95) +
  theme(panel.background = element_rect(fill = 'black'),
        panel.grid.major.x = element_line(color = 'yellow'),
        panel.grid.minor.x = element_line(color = 'white'))
```

We don't have to go into such fine detail, we can use "macro" themes to make it easier to change the backgrounds.  

```{r, iris_27}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, 
             mapping = aes(color = Species)) +
  scale_color_discrete(l = 95) +
  theme_dark()
```

I personally like the black-white theme and I use darker colors.  

```{r, iris_28}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, 
             mapping = aes(color = Species)) +
  scale_color_discrete(l = 45) +
  theme_bw()
```

To move the legend we need another call to `theme()`.  

```{r, iris_29}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, 
             mapping = aes(color = Species)) +
  scale_color_discrete(l = 45) +
  theme_bw() +
  theme(legend.position = 'top')
```

Use the brewer color palettes.  

```{r, iris_30}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3.5, 
             mapping = aes(color = Species)) +
  scale_color_brewer(palette = 'Dark2') +
  theme_bw() +
  theme(legend.position = 'top')
```

There are limited number of colors in the brewer qualitative palettes.  

```{r, iris_31}
set.seed(2130)
iris %>% 
  mutate(random_letters = sample(LETTERS, size = n(), replace = TRUE)) %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = random_letters),
             size = 3.5) +
  scale_color_brewer(palette = 'Set1') +
  theme_bw()
```

Use color blind friendly palettes from `ggthemes`.  

```{r, iris_32}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species),
             size = 3.5) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = 'top')
```

Another palette from `ggthemes`.  

```{r, iris_33}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species),
             size = 3.5) +
  ggthemes::scale_color_calc()+
  theme_bw() +
  theme(legend.position = 'top')
```

Both have very limited number of values.  

```{r, iris_34}
set.seed(2130)
iris %>% 
  mutate(random_letters = sample(LETTERS, size = n(), replace = TRUE)) %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = random_letters),
             size = 3.5) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

Excel colors.  

```{r, iris_35}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species),
             size = 3.5) +
  ggthemes::scale_color_excel_new(theme = 'Office Theme') +
  theme_bw() +
  theme(legend.position = 'top')
```

Excel theme.  

```{r}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species),
             size = 3.5) +
  ggthemes::theme_excel()
```

Economist has a theme.  

```{r}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species),
             size = 3.5) +
  ggthemes::theme_economist()
```


```{r}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species),
             size = 3.5) +
  ggthemes::theme_fivethirtyeight()
```


### Trends per group

Include separate trends per category or Grouping value.  

```{r, iris_36}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species),
             size = 3.5) +
  geom_smooth(formula = y ~ x,
              method = lm,
              mapping = aes(color = Species)) +
  ggthemes::scale_color_colorblind() +
  theme_bw()
```

We can change the ribbon color by the `fill` aesthetic.  

```{r, iris_37}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 3.5) +
  geom_smooth(formula = y ~ x,
              method = lm,
              mapping = aes(color = Species,
                            fill = Species)) +
  ggthemes::scale_color_colorblind() +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```

The linetype is like marker shape.  

```{r, iris_38}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 3.5) +
  geom_smooth(formula = y ~ x,
              method = lm,
              mapping = aes(color = Species,
                            fill = Species,
                            linetype = Species)) +
  ggthemes::scale_color_colorblind() +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```

We can include the UNGROUPED linear trend in this figure.  

```{r, iris_39}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_smooth(formula = y ~ x,
              method = lm,
              color = 'red', fill = 'red') +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 3.5) +
  geom_smooth(formula = y ~ x,
              method = lm,
              mapping = aes(color = Species,
                            fill = Species,
                            linetype = Species)) +
  ggthemes::scale_color_colorblind() +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```

Confidence intervals make plots busy...so we can turn them off.  

```{r, iris_40}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_smooth(formula = y ~ x,
              method = lm,
              color = 'red', fill = 'red',
              se = FALSE,
              size = 1.25) +
  geom_point(mapping = aes(color = Species,
                           shape = Species),
             size = 3.5) +
  geom_smooth(formula = y ~ x,
              method = lm,
              mapping = aes(color = Species,
                            fill = Species,
                            linetype = Species),
              se = FALSE,
              size = 1.25) +
  ggthemes::scale_color_colorblind() +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```

I **never** recommend showing a smoothing line WITHOUT markers.  

```{r, iris_41}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_smooth(formula = y ~ x,
              method = lm,
              color = 'red', fill = 'red',
              se = FALSE,
              size = 1.25) +
  # geom_point(mapping = aes(color = Species,
  #                          shape = Species),
  #            size = 3.5) +
  geom_smooth(formula = y ~ x,
              method = lm,
              mapping = aes(color = Species,
                            fill = Species,
                            linetype = Species),
              se = FALSE,
              size = 1.25) +
  ggthemes::scale_color_colorblind() +
  ggthemes::scale_fill_colorblind() +
  theme_bw()
```

### Continuous colors

We can have 3 variables on a scatter plot where the 3rd variable is continuous and we use color to show the values of that continuous variable.  

```{r, iris_42}
iris %>% summary()
```


Color based on `Petal.Length`.  

```{r, iris_43}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 3.5) +
  theme_bw()
```

We can use the brewer library's distiller function to create other continuous color palletes.  

```{r, iris_44}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 3.5) +
  scale_color_distiller(palette = 'PuBuGn') +
  theme_bw()
```

I prefer the viridis library for sequential continuous color scales.  

```{r, iris_45}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 3.5) +
  scale_color_viridis_c() +
  theme_bw()
```

There are multiple viridis options.  

```{r, iris_46}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 3.5) +
  scale_color_viridis_c(option = 'turbo') +
  theme_bw()
```

We can bin or categorize a continuous variable.  

```{r, iris_47}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 3.5) +
  scale_color_viridis_b(option = 'magma', direction = -1) +
  theme_bw()
```

### 4 variables

We can use shape for categorical combined with color for continuous.  

```{r, iris_48}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length,
                           shape = Species),
             size = 3.5) +
  scale_color_viridis_b(option = 'magma', direction = -1) +
  theme_bw()
```


Alternatively we can use facets for categorical variables.  

```{r, iris_49}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 3.5) +
  facet_wrap(~Species) +
  scale_color_viridis_b(option = 'magma', direction = -1) +
  theme_bw()
```

Change the facet strip background.  

```{r, iris_50}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length),
             size = 3.5) +
  facet_wrap(~Species) +
  scale_color_viridis_b(option = 'magma', direction = -1) +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(size = 12))
```

We can also use color for a categorical. In-line define a boolean for petal being larger than the median petal length.  

```{r, iris_51}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length > median(Petal.Length),
                           shape = Species),
             size = 3.5) +
  facet_wrap(~Species) +
  scale_color_viridis_d(option = 'viridis') +
  scale_shape_discrete(guide = 'none') +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(size = 12),
        legend.position = 'top',
        panel.grid = element_line(color = 'black'))
```


We can use size and transparency for continuous variables. So we can show 4 continuous variables on the same scatter plot.  

```{r, iris_52}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length,
                           size = Petal.Width)) +
  scale_color_viridis_c(option = 'turbo') +
  theme_bw()
```

### 5 variables

We can show 4 continuous variables and 1 categorical on a single scatter plot.  

```{r, iris_53}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length,
                           size = Petal.Width,
                           shape = Species)) +
  scale_color_viridis_c(option = 'turbo') +
  theme_bw()
```

Alternatively we can use facets for the categorical.  

```{r, iris_54}
iris %>% 
  ggplot(mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(mapping = aes(color = Petal.Length,
                           size = Petal.Width)) +
  facet_wrap(~Species) +
  scale_color_viridis_c(option = 'turbo') +
  theme_bw()
```

But if we look at the petal variables...  

```{r, iris_55}
iris %>% 
  ggplot(mapping = aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(size = 3.5) +
  theme_bw()
```

## What are looking at with Scatter plots?  


We are trying to find **linear relationships** or **correlations** between PAIRS of variables.  

Can we get a high level overview of the **correlation** between **ALL** pairs of variables?  

To use corrplot we need to calculate the **correlation matrix**.  

```{r, cc_1}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor()
```

Instead of printing or displaying the correlation plot, let's visualize it with `corrplot()`.  

```{r, cc_2}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot()
```


We can override the default arguments to show just the upper triangular portion.  

```{r, cc_3}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(type = 'upper')
```

We can show numbers instead of circles.  

```{r, cc_4}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(type = 'upper', method = 'number')
```

Or can use a mixture of circles and numbers.  

```{r, cc_5}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```

We can use other methods such as squares.  

```{r, cc_6}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(type = 'upper', method = 'square')
```

The pie chart can be thrown in...  

```{r, cc_7}
iris %>% 
  purrr::keep(is.numeric) %>% 
  cor() %>% 
  corrplot::corrplot(type = 'upper', method = 'pie')
```

### Tidyverse correlation plots

We will use `corrr`.  

```{r, load_corrr_package}
library(corrr)
```

```{r, dd_1}
iris %>% 
  purrr::keep(is.numeric) %>% 
  corrr::correlate()
```

I like to include 1's along the main diagonal.  

```{r, dd_2}
iris %>% 
  purrr::keep(is.numeric) %>% 
  corrr::correlate(diagonal = 1, quiet = TRUE)
```

The `corrr` package has built in plotting functions dedicated to correlation objects.  

```{r, dd_3}
iris_corrr <- iris %>% 
  purrr::keep(is.numeric) %>% 
  corrr::correlate(diagonal = 1, quiet = TRUE)
```

The default plot is created by `rplot()`.  

```{r, dd_4}
iris_corrr %>% rplot()
```

It also has a slick network diagram to show correlations.  

```{r, dd_5}
iris_corrr %>% network_plot()
```

So how can we group?  

```{r, dd_6}
iris_corrr %>% 
  corrr::stretch()
```

We can now manually create our own correlation plot.  

```{r, dd_7}
iris_corrr %>% 
  corrr::stretch() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = r,
                           size = r)) +
  coord_equal() +
  scale_color_gradient2(low = 'red', mid = 'white', high = 'blue',
                        midpoint = 0,
                        limits = c(-1, 1)) +
  scale_size_continuous(guide = 'none',
                        limits = c(-1, 1)) +
  labs(x = '', y ='') +
  theme_bw()
```

To create a grouped object we need to use a grouping operation.  

```{r, dd_8}
iris %>% 
  group_by(Species) %>% 
  group_modify( ~stretch( correlate(., diagonal = 1, quiet = TRUE) ) ) %>% 
  ungroup() %>% 
  ggplot(mapping = aes(x = x, y = y)) +
  geom_point(mapping = aes(color = r,
                           size = r)) +
  coord_equal() +
  facet_wrap(~Species) +
  scale_color_gradient2(low = 'red', mid = 'white', high = 'blue',
                        midpoint = 0,
                        limits = c(-1, 1)) +
  scale_size_continuous(guide = 'none',
                        limits = c(-1, 1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```



